A Stopped Negative Binomial Distribution

### 2016-03-10 stat-and-prob-letters/snb-paper.tex
Made the easy revisions suggested by Dan.

\title{A Stopped Negative Binomial Distribution}
\tnotetext[mytitlenote]{This research was supported by grants R01CA131301, 
R01CA157749, R01CA148996, R01CA168733, and PC50CA196530 awarded by the 
National Cancer Institute, and support from the Yale Comprehensive Cancer 
Center.}

@@

\begin{abstract}
This paper introduces a novel discrete distribution suggested by curtailed
sampling rules common in early-stage clinical trials. We derive the
distribution of the smallest number of independent and identically
distributed Bernoulli trials needed in order to observe either $s$ successes 
or $t$ failures. The closed-form expression for the distribution as well as 
its characteristics and properties are explored.
\end{abstract}

@@

Consider a prototypical Phase II single-arm clinical trial. 12 patients
are enrolled and treated. If at least 2 of those patients respond
favorably, then the trial will proceed to the next stage. If fewer than 
two respond then the trial would have been terminated.

The maximum sample size is 12 but the number of patients necessary to
reach any endpoint could be less. Our goal is to describe the distribution of
the enrollment size. This distribution will be used for planning purposes. 
If all 12 patients are enrolled at once, as in the classic
design, then the sample size is 12. However, in most clinical trials, the
patients are enrolled sequentially, often with one patient's outcome realized
before the next one enters the trial. In the present example, observing two
successful patients allows us reach one endpoint so the sample required
could be as small as two - 12 might not be necessary. Similarly 11
observed treatment failures also ends the stage. This sampling mechanism, in
which the experiment ends as soon as any of the endpoints is reached, is
call {\em curtailed sampling}. Under curtailed sampling the range of the
sample size is between two and 12.

@@

Tab.~\ref{tab:snb} summarizes the distribution and 
some of its characteristics.
The rest of this paper derives these results
and explores properties of the distribution.
The next section introduces our notation and basic results
including the density of the distribution along with a description of
it's relation to other distributions. Sections 2 derives the distribution
based on a defined Bernoulli process and give some basic properties.
Section 3 provides a connection to the Binomial tail probability.
Section 4 derives the posterior distribution using a Beta prior.
Section 5 provides a brief discussion on the use of the distribution
in clinical trials along with future avenues for generalization.

\section{Distributional Result}
\label{notation.section}

Let $\,b_1, b_2, \ldots \,$ denote a sequence of independent, identically
distributed, Bernoulli random variables with $\mathbb{P}[b_i=1]=p$ and
$\mathbb{P}[b_i = 0] = 1-p$, for
probability parameter $0\leq p \leq 1$. In the clinical trial setting
$\,b_i = 1$ corresponds to a successful outcome.  Let $s$ and $t$ be
positive integers.  Define the stopped negative binomial (SNB) random
variable $Y$ as the smallest
integer value such that $\,\{b_1, \ldots , b_Y\}\,$ contains {\em either}
$\,s\,$ successes {\em or} $\,t\,$ failures. That is, either
$\sum_i^Y b_i = s$ or $\sum_i^Y 1-b_i = t$.

The distribution of $\,Y\,$ has support on integer values in the range
\begin{equation*}               
     \min(s,t) \leq \; Y \;\leq s+t-1  \label{range.y.eq}
\end{equation*}
and it is distributed as
\begin{equation} \label{eqn:pmf}
\mathbb{E}\  I_{\{Y=k\}} = S(k, p, s) \ I_{\{s \leq k\}} + 
  S(k, 1-p, t) \ I_{\{ t \leq k \}}
\end{equation}
where $I_{\{f\}}$ is one if $f$ is true and zero otherwise and
\begin{equation} \label{eqn:N}
S(k, p, s) = {k-1 \choose s-1} p^s (1-p)^{k-s} 
\end{equation}
$S$ is the negative binomial probability

To prove the result in Eqn.~\ref{eqn:pmf} consider the
process $\mathbf{X} = \left\{X(k) : k = 0,1,... \right\}$
with $X(0)=0$ and
\begin{equation*} \label{eqn:proc}
X_{k+1} = X_k + b_{k+1} \ I_{\{ k-t < X_k < s\}}.
\end{equation*}
The process can be conceptualized as a series of coin
flips that stops when either $s$ successes or $t$ failures are reached.
At each step a coin is flipped. If it is heads, the process advances
one diagonally in the
positive horizontal and vertical direction.
Otherwise, it advances in the positive horizontal direction only. The
process stops when either $X_k = s$ or $X_k = k-t$.

\begin{prop}
The distribution of the stopping time
$\argminl_k \left[X_k \geq s \cup X_k \leq k-t \right]$
is equal to Eqn.~\ref{eqn:pmf}.
\end{prop}
\begin{proof}
%The proof will proceed in two parts. First, a combinatorial justification 
%will be given for the probability mass value on each element of the support. 
%Second, it will be shown that the sum of the masses over the support sums to 
%one.

The probability a given realization of $\mathbf{X}$ reaches $s$ at
the $k$th outcome is the probability that, at time $k-1$ there are $s-1$
successful outcomes and $k-s$ unsuccessful outcomes multiplied by
the probability of a
success at time $k$. 
\begin{equation} \label{stop_s}
S(k, p, s) = {k-1 \choose s-1} p^s (1-p)^{k-s}
\end{equation}
The probability a given realization reaches $k-t$
is the probability that, at outcome $k-1$ there are $k-t$ successful outcomes
and $t-1$ unsuccessful outcomes multiplied by the probability of an
unsuccessful outcome at time $k$.  
\begin{equation} \label{stop_t}
S'(k, p, t) = {k-1 \choose k-t} p^{k-t} (1-p)^t
\end{equation}
It can be shown that $S(k, p, s) = S'(k, 1-p, s)$ by realizing that 
${k-1 \choose k-s} = {k-1 \choose s-1}$.

To show that sum of Eqn. \ref{stop_s} and \ref{stop_t} sum to one
over their support 
\begin{align} \label{eqn:sum_proof}
R &= \sum_{k=s}^{s+t-1} S(k, p, s) + \sum_{k=t}^{s+t-1} S(k, 1-p, t) \\
  &= \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s (1-p)^{k-s} + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} (1-p)^t
\end{align}
substitute $i=k-s$ in the first summation and $j=k-t$ in the second.
Then $R$ can be written
as the cumulative distribution function of two
negative binomial distributions
\begin{equation} \label{eqn:transformed_sum}
R = \sum_{i=0}^{t-1} {i+s-1 \choose i} p^s (1-p)^i \; + \;
\sum_{j=0}^{s-1} {j+t-1 \choose j} p^j (1-p)^t.
\end{equation}

Let $\mathcal{I}_p(s, t)$ be the {\em regularized incomplete beta function}
and recall this function satisfies 
$\mathcal{I}_p(s, t) = 1-\mathcal{I}_{1-p}(t, s)$ \citep{Abramowitz1964}.
\begin{align*}
R = \sum_{i=0}^{t-1} &{i+s-1 \choose i} p^s (1-p)^i +
\sum_{j=0}^{s-1}  {j+t-1 \choose j} p^j  (1-p)^t \\
   &= 1-\mathcal{I}_p(s, t) + 1 - \mathcal{I}_{1-p}(t, s) \\
   &= 1. 
\end{align*}
\end{proof}

@@

The SNB is related to the negative 
binomial distribution. Specifically, if $t$ is large then the $Y-s$ has a 
negative binomial distribution with
\begin{equation*}                                    %   (1)   
\mathbb{P}[Y=s+j]        \label{nb1.eq}          
  = {{s+j-1}\choose{s-1}} p^s (1-p)^j
\end{equation*}
for $\,j=0, 1,\ldots\,$. A similar statement can be made when $s$ is large
and $t$ is small. As a result, with proper parameter choice, the SNB
can mimic other probability distributions in a manner similar to 
that described in \cite{Peizer1968} and \cite{Best1974}. Examples are
shown in in Fig.~\ref{shapes.fig}.

For the special case of $\,s=t,$ the distribution of $\,Y\,$ is the
riff-shuffle, or minimum negative binomial distribution~\citep{Uppuluri1970}.
Similar derivations of the closely-related maximum negative binomial 
discrete distributions also appear in~\cite{Zhang2000}
and \cite{Zelterman2005}.
The maximum negative binomial is the smallest number of outcomes necessary to 
observe at least $c$ successes {\em and} $c$ failures. The SNB is the 
number of coin flips to observe {\em either} $s$ heads or $t$ tails.

\section{Connection Between $S$ and the Binomial Tail Probability}

\begin{prop} \label{binomial_tail}
Let $Y$ be distributed as SNB($p$, $s$, $t$) and let 
$B$ be distributed Binomial with size $n=s+t-1$ and success probability
$p$. Then
\begin{equation}
\mathbb{P}[B \geq s] = \mathbb{P} [Y \leq n\ |\ \text{\#success} = s].
\end{equation}
That is, the probability that the number of successes is at least $s$
in the Binomial model is the same that the trial stops with $s$ 
successes in the SNB model.
\end{prop}
\begin{proof}
The Binomial tail probability is
\begin{align*}
\mathbb{P}[B \geq s] &= \sum_{k=s}^{s+t-1} {n \choose k} p^k (1-p)^{n-k} \\
  &= 1 - \sum_{k=0}^{s-1} {n \choose k} p^k (1-p)^{n-k} \\
  & 1 - \mathcal{I}_{1-p}.
\end{align*}
Next, consider the SNB probability
\begin{align*}
\mathbb{P} [Y \leq n\ |\ \text{\#success}=s] &= \sum_{k=s}^{s+t-1} S(k, p, s)\\
 &= {k-1 \choose s-1} p^s (1-p)^{k-s}.
\end{align*}
Let $i=k-s$ then using the fact that ${i+s-1 \choose s-1} = {i+s-1 \choose i}$
the summation can be rewritten as
\begin{align}
\mathbb{P} [Y \leq n\ |\ \text{\#success} = s] &= \sum_{i=0}^{t-1} 
  {i+s-1 \choose i} p^s (1-p)^i\\
  &= 1 - \mathcal{I}_{1-p}(t, s).
\end{align}
\end{proof}

This result is visualized in Fig.~\ref{fig:snb_bin_compare} for the case
where $s=2$, $t=12$, and $p=0.2$. The probability masses represented in
red are equal as are the masses in teal. The probability that $s$
successes are reached in the SNB process is the same as the binomial 
probability of at least two successes. Likewise, the probability that $t$ 
failures are reached in the SNB process is the same as the binomial
probability of zero or one successes.

@@

\begin{prop} Let $Y$ be distributed SNB with parameters $p$, $s$, and $t$.
Then the moment generating function (MGF) of $Y$ is
\begin{equation} \label{eqn:mgf}
\mathbb{E}~e^{xY} = \left(\frac{p e^x}{1 - qe^x}\right)^s 
  \mathcal{I}_{1-qe^x} (s, t) + \left(\frac{qe^x}{1-pe^x}\right)^t 
  \mathcal{I}_{1-pe^x}(t, s)
\end{equation}
for $q = 1-p$ when $x \leq -\log(q)$ and $x \leq -\log(p)$.
\end{prop}
\begin{proof}
By definition, the MGF of the SNB is:
\begin{equation*}
\mathbb{E}~e^{xY} = \sum_{k=s}^{s+t-1} {k-1 \choose k-s} p^s q^{k-s} e^{kx} 
  + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} q^t e^{kx}
\end{equation*}
and can be rewritten as:
\begin{equation} \label{eqn:first_sum}
\mathbb{E}~e^{xY} = \sum_{k=s}^{s+t-1}{k-1 \choose k-s} (pe)^{sx} (qe^x)^{k-s} 
  + \sum_{k=t}^{s+t-1}{k-1 \choose k-t} (qe^x)^t (pe^x)^{k-t}.
\end{equation}
Taking the first summation in Equation \ref{eqn:first_sum}:
\begin{align*}
\sum_{k=s}^{s+t-1}{k-1 \choose k-s} (pe)^{sx} (qe^x)^{k-s} &= 
  \left(\frac{pe^x}{1 - qe^x}\right)^s \ \ \sum_{k=s}^{s+t-1} {k-1 \choose k-s} 
    (qe^x)^{k-s} (1-qe^x)^s \\
  &= \left(\frac{pe^x}{1 - qe^x}\right)^s \mathcal{I}_{qe^x}(s, t).
\end{align*}
Since the incomplete beta function has support on zero to one
$qe^x \leq 1$. This implies that $x \leq -\log(q)$.

@@

\begin{prop}
The posterior PMF of the Stopped Negative Binomial distribution with a Beta($\alpha$, $\beta$) prior is:
\begin{align} \label{eqn:posterior}
f(k | s, t, \alpha, \beta) &= 
  {k-1 \choose s-1} \frac{B\left(\alpha+s, k-s+\beta \right)}{B(\alpha, \beta)} 
    \ I_{\{s \leq k \leq s+k-1\}} \ + \nonumber \\
  & {k-1 \choose k-t} 
    \frac{B\left(\alpha + k - t, t+\beta\right)}{B(\alpha, \beta)} 
    \ I_{\{t \leq k \leq s+k-1\}}
\end{align}
where $B$ is the Beta function.
\end{prop}
\begin{proof}
For notational simplicity, assume that $s,t \leq k \leq s+t-1$. When this is not the case appropriate terms should be removed as dictated by the indicator functions.
\begin{align*}
f(k | s, t, \alpha, \beta) = \frac{1}{B(\alpha, \beta)} & \int_0^1 {k-1 \choose s-1} p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} + \\
 & {k-1 \choose k-t} p^{k-t+\alpha-1}\left(1-p\right)^{t+b-1} dp \\
= \frac{1}{B(\alpha, \beta)}  {k-1 \choose s-1} & \int_0^1  p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} dp + \\
 & \frac{1}{B(\alpha, \beta)} {k-1 \choose k-t} \int_0^1  p^{k-t+\alpha-1}\left(1-p\right)^{t+b-1} dp
\end{align*}
The result follows by applying the definition of the Beta function to the integral terms.
\end{proof}

@@

We have presented a new discrete distribution by curtailed sampling rules common in early-stage clinical trials, which we refer to as the Stopped Negative Binomial distribution. The distribution models the stopping time of a sequential trial where the trial is stopped when a number of events are accumulated. The
posterior distribution was derived for the case when the event probability $p$
has a Beta distribution. Using a trial description from
\url{clinicaltrials.gov} we showed how the SNB is an integral part of trial
monitoring, and post-hoc analysis and can be used in a framework alternative
to the Simon two-stage optimal design while providing better estimates of the
event probability along with quantifying the uncertainty associated with the
estimates. As a result, we were able to show fewer patient enrollments and
achieve the same estimates, when added in a hypothetical, but representative,
clinical trial.

Current work focuses on the generalization of the distribution and the application in other areas of clinical trials. Adverse outcomes in particular are another area that could greatly benefit both from the presented distribution and its framework. For these trials monitoring may need to be performed not only on the outcome of an intervention but also on the safety of the trial. This especially
true in areas like late-stage cancer treatments where the treatment is harsh
and patients may be forced to drop out as a result of the side-effects of the
treatment; a matter independent of the outcome. Other application areas
include providing designs that allows clinicians to balance uncertainty and
success probability with a minimum number of patient enrollees.

>>>>

\title{A Stopped Negative Binomial Distribution}
\tnotetext[mytitlenote]{This research was supported by grants R01CA131301, 
R01CA157749, R01CA148996, R01CA168733, and PC50CA196530 awarded by the 
National Cancer Institute along with support from the Yale Comprehensive Cancer 
Center and the Yale Center for Outcomes Research. We would also like 
to thank Rick Landin at Ignyta Inc. for his suggestions.}

@@

\begin{abstract}
We introduce a discrete distribution suggested by curtailed
sampling rules common in early-stage clinical trials. We derive the
distribution of the smallest number of independent and identically
distributed Bernoulli trials needed to observe either $s$ successes 
or $t$ failures. The closed-form expression for the distribution as well as 
its characteristics are derived and properties of the distribution are explored.
\end{abstract}

@@

Consider a prototypical Phase II single-arm clinical trial in which 12 patients
are enrolled and treated. If two or more respond then we reject the null
hypothesis that the treatment is not effective and the trial proceeds to the 
next stage. If fewer that two respond then the null is not rejected and the 
trial is terminated.

If all 12 patients are enrolled at once, as in the classic
design, then the sample size is 12. However, as in most clinical trials, the
patients are enrolled sequentially, often with one patient's outcome realized
before the next one enters the trial. In the present example, observing two
successful patients allows us reach one endpoint so the sample required
could be as small as two. Similarly 11
observed treatment failures also ends the stage. This sampling mechanism, in
which the experiment ends as soon as any of the endpoints is reached, is
call {\em curtailed sampling}. Under curtailed sampling the range of the
sample size is between two and 12.

@@

In this work we derive the distribution of the number of enrollees needed
for either $s$ successes or $t$ failures. We refer to this distribution
as the Stopped Negative Binomial (SNB) and some of its characteristics are
summarized in Tab.~\ref{tab:snb}.
The rest of this paper derives these results
and explores properties of the distribution.
The next section introduces our notation and basic results
including the density of the distribution along with a description of
it's relation to other distributions. Section 2 derives the distribution
based on a defined Bernoulli process and gives some basic properties.
Section 3 provides a connection to the Binomial tail probability.
Section 4 derives the posterior distribution using a Beta prior.
Section 5 provides a brief discussion on the use of the distribution
in clinical trials along with future avenues for generalization.

\section{Probability Mass Function}
\label{notation.section}

Let $\,b_1, b_2, \ldots \,$ denote a sequence of independent, identically
distributed, Bernoulli random variables with $\mathbb{P}[b_i=1]=p$ and
$\mathbb{P}[b_i = 0] = 1-p$, for
probability parameter $0\leq p \leq 1$. In the clinical trial setting
$\,b_i = 1$ corresponds to a successful outcome.  Let $s$ and $t$ be
positive integers.  Define the SNB random
variable $Y$ as the smallest
integer value such that $\,\{b_1, \ldots , b_Y\}\,$ contains {\em either}
$\,s\,$ successes {\em or} $\,t\,$ failures. That is, the SNB distribution
of $Y$ is the smallest integer such that either
$\sum_i^Y b_i = s$ or $\sum_i^Y 1-b_i = t$.

The distribution of $\,Y\,$ has support on integer values in the range
\begin{equation*}               
     \min(s,t) \leq \; Y \;\leq s+t-1  \label{range.y.eq}
\end{equation*}
and it is distributed as
\begin{equation} \label{eqn:pmf}
\mathbb{E}\  I_{\{Y=k\}} = S(k, p, s) \ I_{\{s \leq k\}} + 
  S(k, 1-p, t) \ I_{\{ t \leq k \}}
\end{equation}
where $I_{\{f\}}$ is one if $f$ is true and zero otherwise and
\begin{equation} \label{eqn:N}
S(k, p, s) = {k-1 \choose s-1} p^s (1-p)^{k-s} 
\end{equation}
is the negative binomial probability

To prove (\ref{eqn:pmf}), consider the
process $\mathbf{X} = \left\{X(k) : k = 0,1,... \right\}$
with $X(0)=0$ and
\begin{equation*} \label{eqn:proc}
X_{k+1} = X_k + b_{k+1} \ I_{\{ k-t < X_k < s\}}.
\end{equation*}
At each step a patient's outcome is measured. If it is success, the process 
advances one diagonally in the
positive horizontal and vertical direction.
Otherwise, it advances in the positive horizontal direction only. The
process continues until either $X_k = s$ or $X_k = k-t$.

\begin{prop}
The distribution of the stopping time
$\argminl_k \left[X_k \geq s \cup X_k \leq k-t \right]$
is given at (\ref{eqn:pmf}).
\end{prop}
\begin{proof}
%The proof will proceed in two parts. First, a combinatorial justification 
%will be given for the probability mass value on each element of the support. 
%Second, it will be shown that the sum of the masses over the support sums to 
%one.

The probability a given realization of $\mathbf{X}$ reaches $s$ at
the $k$th outcome is the probability that, at time $k-1$ there are $s-1$
successful outcomes and $k-s$ unsuccessful outcomes multiplied by
the probability of a success at time $k$. This expression is given
in (\ref{eqn:N}). 
Similarly, probability a given realization reaches $k-t$
is the probability that, at outcome $k-1$ there are $k-t$ successful outcomes
and $t-1$ unsuccessful outcomes multiplied by the probability of an
unsuccessful outcome at time $k$.

Next, define
\begin{equation} \label{stop_t}
S'(k, p, t) = {k-1 \choose k-t} p^{k-t} (1-p)^t
\end{equation}
and notice that $S(k, p, s) = S'(k, 1-p, s)$ by writing
${k-1 \choose k-s} = {k-1 \choose s-1}$.

To show that (\ref{eqn:N}) and (\ref{stop_t}) sum to one
over their support, let
\begin{align} \label{eqn:sum_proof}
R &= \sum_{k=s}^{s+t-1} S(k, p, s) + \sum_{k=t}^{s+t-1} S(k, 1-p, t) \\
  &= \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s (1-p)^{k-s} + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} (1-p)^t
\end{align}
where we substitute $i=k-s$ in the first summation and $j=k-t$ in the second.

Then $R$ can be written as the cumulative distribution function of two
negative binomial distributions:
\begin{equation} \label{eqn:transformed_sum}
R = \sum_{i=0}^{t-1} {i+s-1 \choose i} p^s (1-p)^i \; + \;
\sum_{j=0}^{s-1} {j+t-1 \choose j} p^j (1-p)^t.
\end{equation}

Let $\mathcal{I}_p(s, t)$ be the {\em regularized incomplete beta function} 
\citep{Olver2010} and recall this function satisfies 
$\mathcal{I}_p(s, t) = 1-\mathcal{I}_{1-p}(t, s)$ \citep{Abramowitz1964}.
\begin{align*}
R = \sum_{i=0}^{t-1} &{i+s-1 \choose i} p^s (1-p)^i +
\sum_{j=0}^{s-1}  {j+t-1 \choose j} p^j  (1-p)^t \\
   &= 1-\mathcal{I}_p(s, t) + 1 - \mathcal{I}_{1-p}(t, s) \\
   &= 1. 
\end{align*}
This completes the proof that (\ref{eqn:pmf}) is a valid probability mass
function.
\end{proof}

@@

The SNB is a generalization of the negative 
binomial distribution. If $t$ is large then the $Y-s$ has a 
negative binomial distribution with
\begin{equation*}                                    %   (1)   
\mathbb{P}[Y=s+j]        \label{nb1.eq}          
  = {{s+j-1}\choose{s-1}} p^s (1-p)^j
\end{equation*}
for $\,j=0, 1,\ldots\,$. A similar statement can be made when $s$ is large
and $t$ is small. As a result, with proper parameter choice, the SNB
can mimic other probability distributions in a manner similar to 
those described in \cite{Peizer1968} and \cite{Best1974}. Examples are
shown in in Fig.~\ref{shapes.fig}.

For the special case of $\,s=t,$ the distribution of $\,Y\,$ is the
riff-shuffle, or minimum negative binomial distribution~\citep{Uppuluri1970}.
Similar derivations of the closely-related maximum negative binomial 
discrete distributions also appear in~\cite{Zhang2000}
and \cite{Zelterman2005}.
The maximum negative binomial is the smallest number of outcomes necessary to 
observe at least $s$ successes {\em and} $s$ failures. The SNB is the 
number of coin flips to observe {\em either} $s$ heads or $t$ tails.

\section{Connection Between the SNB and the Binomial Tail Probability}

\begin{prop} \label{binomial_tail}
Let $Y$ be distributed as SNB($p$, $s$, $t$) and let 
$B$ be distributed Binomial with size $n=s+t-1$ and success probability
$p$. Then
\begin{equation}
\mathbb{P}[B \geq s] = \mathbb{P} [Y \leq n\ |\ \text{\#success} = s].
\end{equation}
That is, the probability that the number of successes is at least $s$
in the Binomial model is the same that the trial stops with $s$ 
successes in the SNB model.
\end{prop}
\begin{proof}
The Binomial tail probability is
\begin{align*}
\mathbb{P}[B \geq s] &= \sum_{k=s}^{s+t-1} {n \choose k} p^k (1-p)^{n-k} \\
  &= 1 - \sum_{k=0}^{s-1} {n \choose k} p^k (1-p)^{n-k} \\
  & 1 - \mathcal{I}_{1-p}.
\end{align*}
The corresponding SNB probability is
\begin{align*}
\mathbb{P} [Y \leq n\ |\ \text{\#success}=s] &= \sum_{k=s}^{s+t-1} S(k, p, s)\\
 &= {k-1 \choose s-1} p^s (1-p)^{k-s}.
\end{align*}
Let $i=k-s$. Use ${i+s-1 \choose s-1} = {i+s-1 \choose i}$
so the last summation can be rewritten as
\begin{align}
\mathbb{P} [Y \leq n\ |\ \text{\#success} = s] &= \sum_{i=0}^{t-1} 
  {i+s-1 \choose i} p^s (1-p)^i\\
  &= 1 - \mathcal{I}_{1-p}(t, s)
\end{align}
completing the proof.
\end{proof}

Fig.~\ref{fig:snb_bin_compare} shows the case
where $s=2$, $t=12$, and $p=0.2$. The probability masses represented in
red are equal as are the masses in teal. The probability that $s$
successes are reached in the SNB process is the same as the binomial 
probability of at least two successes. Likewise, the probability that $t$ 
failures are reached in the SNB process is the same as the binomial
probability of zero or one successes.

@@

\begin{prop} Let $Y$ be distributed SNB with parameters $p$, $s$, and $t$.
Then the moment generating function (MGF) of $Y$ is
\begin{equation} \label{eqn:mgf}
\mathbb{E}~e^{xY} = \left(\frac{p e^x}{1 - qe^x}\right)^s 
  \mathcal{I}_{1-qe^x} (s, t) + \left(\frac{qe^x}{1-pe^x}\right)^t 
  \mathcal{I}_{1-pe^x}(t, s)
\end{equation}
for $q = 1-p$ when $x \leq \min \left\{\log(1/p), \log(1/q) \right\}$.
\end{prop}
\begin{proof}
The MGF of the SNB is:
\begin{equation*}
\mathbb{E}~e^{xY} = \sum_{k=s}^{s+t-1} {k-1 \choose k-s} p^s q^{k-s} e^{kx} 
  + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} q^t e^{kx}
\end{equation*}
and can be rewritten as:
\begin{equation} \label{eqn:first_sum}
\mathbb{E}~e^{xY} = \sum_{k=s}^{s+t-1}{k-1 \choose k-s} (pe)^{sx} (qe^x)^{k-s} 
  + \sum_{k=t}^{s+t-1}{k-1 \choose k-t} (qe^x)^t (pe^x)^{k-t}.
\end{equation}
Taking the first summation in Equation \ref{eqn:first_sum}:
\begin{align*}
\sum_{k=s}^{s+t-1}{k-1 \choose k-s} (pe)^{sx} (qe^x)^{k-s} &= 
  \left(\frac{pe^x}{1 - qe^x}\right)^s \ \ \sum_{k=s}^{s+t-1} {k-1 \choose k-s} 
    (qe^x)^{k-s} (1-qe^x)^s \\
  &= \left(\frac{pe^x}{1 - qe^x}\right)^s \mathcal{I}_{qe^x}(s, t).
\end{align*}
Since the incomplete beta function has support on zero to one, we have
$qe^x \leq 1$. This also shows $x \leq -\log(q)$.

@@

Let us consider a Bayesian analysis where there is a Beta prior distribution
on $p$.
\begin{prop}
The posterior PMF of the Stopped Negative Binomial distribution with a Beta($\alpha$, $\beta$) prior is:
\begin{align} \label{eqn:posterior}
f(k | s, t, \alpha, \beta) &= 
  {k-1 \choose s-1} \frac{B\left(\alpha+s, k-s+\beta \right)}{B(\alpha, \beta)} 
    \ I_{\{s \leq k \leq s+k-1\}} \ + \nonumber \\
  & {k-1 \choose k-t} 
    \frac{B\left(\alpha + k - t, t+\beta\right)}{B(\alpha, \beta)} 
    \ I_{\{t \leq k \leq s+k-1\}}.
\end{align}
\end{prop}
\begin{proof}
For notational simplicity, assume that $s,t \leq k \leq s+t-1$. When this is not the case appropriate terms should be removed as dictated by the indicator functions.
\begin{align*}
f(k | s, t, \alpha, \beta) = \frac{1}{B(\alpha, \beta)} & \int_0^1 {k-1 \choose s-1} p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} + \\
 & {k-1 \choose k-t} p^{k-t+\alpha-1}\left(1-p\right)^{t+b-1} dp \\
= \frac{1}{B(\alpha, \beta)}  {k-1 \choose s-1} & \int_0^1  p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} dp + \\
 & \frac{1}{B(\alpha, \beta)} {k-1 \choose k-t} \int_0^1  p^{k-t+\alpha-1}\left(1-p\right)^{t+b-1} dp
\end{align*}
The result follows by applying the definition of the Beta function to the integral terms.
\end{proof}

@@

We have presented a new discrete distribution by curtailed sampling rules 
common in early-stage clinical trials, which we refer to as the Stopped 
Negative Binomial distribution. The distribution models the stopping time of 
a sequential trial where the trial is stopped when a number of events are 
accumulated. A connection between the Binomial tail probability and the SNB
distribution was shown; It's MGF was derived; and
the posterior distribution was derived for the case when the 
event probability $p$ has a Beta distribution.

Current work focuses on two different areas. First, the distribution
is being applied to clinical trial design
where the number of enrollees is smaller than traditional trials.
This scenario is increasingly common in targeted cancer
therapy where treatment may only be appropriate for small subpopulations.
Second, generalizations and extensions of the distribution are being pursued
for applications power analysis. The posterior construction allows us
to quantify the uncertainty both in the response probability as well as the
time until a trial is complete. This suggests new approaches to sample
size calculations as well as trial monitoring.
